Factors affecting health sector expenditure in OECD countries

### 2025-04-05 code/analysis-code/statistical-analysis.qmd
added RF results

@@

# converting all columns to numeric except CountryName, CountryCode, and SeriesCode
train <- train %>%
  mutate(across(-c(CountryName, CountryCode, SeriesCode), as.numeric))

#saving the train dataset as we will be using it in future as well

saveRDS(train, file = here("data", "processed-data", "train-data.rds"))

# converting all columns to numeric except CountryName, CountryCode, and SeriesCode
test <- test %>%
  mutate(across(-c(CountryName, CountryCode, SeriesCode), as.numeric))

#saving the train dataset as we will be using it in future as well

saveRDS(test, file = here("data", "processed-data", "test-data.rds"))
```

@@

```
ADF test (combined) continued....

@@

>>>>

Grouping the variables.

@@

@@

ADF test (combined) continued....

@@

```{r}
# primary regression model without LE

### 2025-04-05 code/eda-code/eda.qmd
added RF results

>>>>

But first, let's save the dataset

```{r}
# converting all columns to numeric except CountryName, CountryCode, and SeriesCode
train <- train %>%
  mutate(across(-c(CountryName, CountryCode, SeriesCode), as.numeric))

#saving the train dataset as we will be using it in future as well

saveRDS(train, file = here("data", "processed-data", "train-data.rds"))

# converting all columns to numeric except CountryName, CountryCode, and SeriesCode
test <- test %>%
  mutate(across(-c(CountryName, CountryCode, SeriesCode), as.numeric))

#saving the train dataset as we will be using it in future as well

saveRDS(test, file = here("data", "processed-data", "test-data.rds"))
```

### 2025-04-05 code/robustness-code/robustness-code.qmd
added RF results

```{r}
# Load required packages
library(purrr)    # For map_df()
library(dplyr)    # For data manipulation
library(timetk)   # For time_series_cv()
library(rsample)  # For analysis() and assessment()
library(here)
library(modelsummary)
library(tidyr)
library(ranger)

@@

testing the results

```{r}
# preliminary regression model

@@

@@

@@

# 1. Define your model formula (as you originally specified)
model_formula <- Current_health_expenditure_per_capita_current_US ~ 
  GDP_growth_annual_ + 
  Income_share_held_by_lowest_20 + 
  Control_of_Corruption_Estimate + 
  Life_expectancy_at_birth_total_years + 
  Unemployment_youth_total__of_total_labor_force_ages_1524_modeled_ILO_estimate + 
  Trade__of_GDP + 
  Foreign_direct_investment_net_inflows_BoP_current_US + 
  Access_to_electricity_rural__of_rural_population +
  Age_dependency_ratio__of_workingage_population

# 2. Create time-series CV splits
cv_splits <- time_series_cv(
  data = train,
  date_var = Year,
  initial = 5,  # 5-year initial window
  assess = 1,   # 1-year test window
  skip = 1,     # Move forward 1 year each split
  slice_limit = 3
)
```

@@

    # Train model
    model <- lm(model_formula, data = train_cv)

    # Get predictions and actuals
    preds <- predict(model, newdata = test_cv)
    actuals <- test_cv$Current_health_expenditure_per_capita_current_US

    # Convert to numeric safely
    preds_num <- suppressWarnings(as.numeric(preds))
    actuals_num <- suppressWarnings(as.numeric(actuals))

    # Calculate RMSE with error handling
    if(length(preds_num) == 0 || length(actuals_num) == 0) {
      rmse_val <- NA_real_
      status <- "Empty predictions/actuals"
    } else if(any(is.na(actuals_num))) {
      valid <- !is.na(actuals_num)
      rmse_val <- sqrt(mean((preds_num[valid] - actuals_num[valid])^2, na.rm = TRUE))
      status <- "NA in actuals"
    } else {
      rmse_val <- sqrt(mean((preds_num - actuals_num)^2, na.rm = TRUE))
      status <- "Success"
    }

@@

# Run the analysis
results <- cv(cv_splits$splits, model_formula)
print(results)
```

```{r}
# View summary of CV results
cv_summary <- results %>%
  summarise(
    Mean_RMSE = mean(RMSE, na.rm = TRUE),
    SD_RMSE = sd(RMSE, na.rm = TRUE),
  )

@@

@@

    # 1. Get target variable from formula
    target_var <- all.vars(model_formula)[1]

    # 2. Check if Year column exists (case-insensitive)
    year_col <- grep("^year$", names(train_cv), ignore.case = TRUE, value = TRUE)

    # 3. Select required variables (including Year if it exists)
    required_vars <- unique(c(all.vars(model_formula), year_col))
    train_cv <- train_cv %>% 
      select(any_of(required_vars)) %>% 
      drop_na(all_of(target_var))

@@

    # 4. Get year information
    get_years <- function(df) {
      if (length(year_col) > 0) {
        list(
          min_year = min(df[[year_col]], na.rm = TRUE),
          max_year = max(df[[year_col]], na.rm = TRUE)
        )
      } else {
        list(min_year = NA_real_, max_year = NA_real_)
      }
    }

@@

    # 5. Calculate mtry
    n_predictors <- length(setdiff(required_vars, c(target_var, year_col)))
    mtry_value <- max(1, min(floor(n_predictors / 3), n_predictors))

    # 6. Train model
    set.seed(123)
    rf_model <- ranger(
      formula = model_formula,
      data = train_cv,
      num.trees = 500,
      mtry = mtry_value,
      importance = 'permutation',
      seed = 123
    )

    # 7. Calculate RMSE
    preds <- predict(rf_model, data = test_cv)$predictions
    actuals <- pull(test_cv, target_var)
    rmse_val <- sqrt(mean((preds - actuals)^2, na.rm = TRUE))

    # 8. Return results
    tibble(
      Train_Years = if (!is.na(train_years$min_year)) {
        paste(train_years$min_year, train_years$max_year, sep = "-")
      } else {NA_character_},
      Test_Year = test_year,
      RMSE = rmse_val,
      Model = "Random Forest",
      Predictors_Used = n_predictors,
      mtry_Used = mtry_value,
      Year_Column_Found = ifelse(length(year_col) > 0, year_col, "Not found")
    )
  })
}
```

@@

@@

# Prepare training data (remove NA in target variable)
target_var <- all.vars(model_formula)[1]
train_processed <- train %>%
  drop_na(all_of(target_var))

# Calculate appropriate mtry
n_predictors <- length(all.vars(model_formula)) - 1
mtry_value <- max(1, min(floor(n_predictors / 3), n_predictors))

# Train Random Forest model
set.seed(123)
rf_model <- ranger(
  formula = model_formula,
  data = train_processed,
  num.trees = 500,
  mtry = mtry_value,
  importance = 'permutation',
  seed = 123
)
```

```{r}
# Ensure test data has same structure as training data
test_processed <- test %>%
  select(any_of(names(train_processed))) %>%  # Keep only columns present in training
  drop_na(all_of(target_var))  # Remove rows with NA in target
```

```{r}
# Generate predictions
test_predictions <- predict(rf_model, data = test_processed)$predictions

# Create results dataframe
test_results <- test_processed %>%
  mutate(Predicted = test_predictions,
         Residual = .data[[target_var]] - Predicted)
```

```{r}
# Calculate metrics
performance_metrics <- test_results %>%
  summarise(
    RMSE = sqrt(mean(Residual^2, na.rm = TRUE)),
    MAE = mean(abs(Residual), na.rm = TRUE),
    R2 = 1 - (sum(Residual^2) / sum((.data[[target_var]] - mean(.data[[target_var]]))^2))
  )

# Print metrics
cat("\nTest Set Performance:\n")
print(performance_metrics)

# View predictions vs actual
head(test_results %>% select(all_of(target_var), Predicted, Residual))
```

>>>>

```{r}
# Loading required packages
library(purrr)    # For map_df()
library(dplyr)    # For data manipulation
library(timetk)   # For time_series_cv()
library(rsample)  # For analysis() and assessment()
library(here)
library(modelsummary)
library(tidyr)
library(ranger)

@@

Here, we will be testing the regression model on out test data

```{r}
# primary regression model on test data

@@

The results from our test dataset is show moderate performance of our model. Check the manuscript for more details on the comparision between train and test results.

@@

Here, we will use cross validation to test the robustness of our model. We will have 5 year initial window and 1 year test window.

@@

# defining the model  
model_formula <- Current_health_expenditure_per_capita_current_US ~ 
  GDP_growth_annual_ + 
  Income_share_held_by_lowest_20 + 
  Control_of_Corruption_Estimate + 
  Life_expectancy_at_birth_total_years + 
  Unemployment_youth_total__of_total_labor_force_ages_1524_modeled_ILO_estimate + 
  Trade__of_GDP + 
  Foreign_direct_investment_net_inflows_BoP_current_US + 
  Access_to_electricity_rural__of_rural_population +
  Age_dependency_ratio__of_workingage_population

#creating time-series CV splits
cv_splits <- time_series_cv(
  data = train,
  date_var = Year,
  initial = 5,  # 5-year initial window
  assess = 1,   # 1-year test window
  skip = 1,     # Move forward 1 year each split
  slice_limit = 3
)
```

@@

    #training model
    model <- lm(model_formula, data = train_cv)

    # Getting predictions and actuals
    preds <- predict(model, newdata = test_cv)
    actuals <- test_cv$Current_health_expenditure_per_capita_current_US

    #converting to numeric 
    preds_num <- suppressWarnings(as.numeric(preds))
    actuals_num <- suppressWarnings(as.numeric(actuals))

    #calculating RMSE
    if(length(preds_num) == 0 || length(actuals_num) == 0) {
      rmse_val <- NA_real_
      status <- "Empty predictions/actuals"
    } else if(any(is.na(actuals_num))) {
      valid <- !is.na(actuals_num)
      rmse_val <- sqrt(mean((preds_num[valid] - actuals_num[valid])^2, na.rm = TRUE))
      status <- "NA in actuals"
    } else {
      rmse_val <- sqrt(mean((preds_num - actuals_num)^2, na.rm = TRUE))
      status <- "Success"
    }

@@

#saving the results
results <- cv(cv_splits$splits, model_formula)
print(results)
```

```{r}
#view summary of CV results
cv_summary <- results %>%
  summarise(
    Mean_RMSE = mean(RMSE, na.rm = TRUE),
    SD_RMSE = sd(RMSE, na.rm = TRUE),
  )

@@

The mean RMSE is 1151 which is simillar to the RMSE found in the original result (RMSE=1213). Therefore CV estimates show that our result is quite robust with very low Standard Deviation of 74.8

We will also test if there is any other model that performs better than OLS regression. We will use Random Forest model.

@@

    # setting the target variables
    target_var <- all.vars(model_formula)[1]

    # checking if Year column exists
    year_col <- grep("^year$", names(train_cv), ignore.case = TRUE, value = TRUE)

    #selecting required variables
    required_vars <- unique(c(all.vars(model_formula), year_col))
    train_cv <- train_cv %>% 
      select(any_of(required_vars)) %>% 
      drop_na(all_of(target_var))

@@

    # getting years
    get_years <- function(df) {
      if (length(year_col) > 0) {
        list(
          min_year = min(df[[year_col]], na.rm = TRUE),
          max_year = max(df[[year_col]], na.rm = TRUE)
        )
      } else {
        list(min_year = NA_real_, max_year = NA_real_)
      }
    }

@@

    #calculate mtry
    n_predictors <- length(setdiff(required_vars, c(target_var, year_col)))
    mtry_value <- max(1, min(floor(n_predictors / 3), n_predictors))

    #train model
    set.seed(123)
    rf_model <- ranger(
      formula = model_formula,
      data = train_cv,
      num.trees = 500,
      mtry = mtry_value,
      importance = 'permutation',
      seed = 123
    )

    #calculate RMSE
    preds <- predict(rf_model, data = test_cv)$predictions
    actuals <- pull(test_cv, target_var)
    rmse_val <- sqrt(mean((preds - actuals)^2, na.rm = TRUE))

    #return results
    tibble(
      Train_Years = if (!is.na(train_years$min_year)) {
        paste(train_years$min_year, train_years$max_year, sep = "-")
      } else {NA_character_},
      Test_Year = test_year,
      RMSE = rmse_val,
      Model = "Random Forest",
      Predictors_Used = n_predictors,
      mtry_Used = mtry_value,
      Year_Column_Found = ifelse(length(year_col) > 0, year_col, "Not found")
    )
  })
}
```

@@

The results of Random Forest are also quite good. They show very low RMSE compared to the OLS Moldel.

@@

#preparing training data
target_var <- all.vars(model_formula)[1]
train_processed <- train %>%
  drop_na(all_of(target_var))

#calculate appropriate mtry
n_predictors <- length(all.vars(model_formula)) - 1
mtry_value <- max(1, min(floor(n_predictors / 3), n_predictors))

#training Random Forest model
set.seed(123)
rf_model <- ranger(
  formula = model_formula,
  data = train_processed,
  num.trees = 500,
  mtry = mtry_value,
  importance = 'permutation',
  seed = 123
)
```

```{r}
#processing test data
test_processed <- test %>%
  select(any_of(names(train_processed))) %>%  # keeping only columns present in training
  drop_na(all_of(target_var))  # Removing rows with NA in target
```

```{r}
# Generating predictions
test_predictions <- predict(rf_model, data = test_processed)$predictions

#creating results dataframe
test_results <- test_processed %>%
  mutate(Predicted = test_predictions,
         Residual = .data[[target_var]] - Predicted)
```

```{r}
#calculating metrics
performance_metrics <- test_results %>%
  summarise(
    RMSE = sqrt(mean(Residual^2, na.rm = TRUE)),
    MAE = mean(abs(Residual), na.rm = TRUE),
    R2 = 1 - (sum(Residual^2) / sum((.data[[target_var]] - mean(.data[[target_var]]))^2))
  )

# printing metrics
cat("\nTest Set Performance:\n")
print(performance_metrics)

However, the results from test dataset are not promising for our Random Forest model. So, we will stick with our OLS model.

```{r}
#viewing the predictions vs actuals
head(test_results %>% select(all_of(target_var), Predicted, Residual))
```

### 2025-04-05 products/manuscript/Manuscript.qmd
added RF results

We can observe that the variables in general show some common pattern between the two datasets. The difference is acceptable as they are based on entirely different time frames. If we observe the model performance, it is evident that the model performance is weaker than obtined in the train dataset. The RMSE has increased by 383 from 1213 to 1596, however this is still 768 lower than the NULL model. The adjusted R-squared in our test dataset has also dropped by 12.3% to 62%

>>>>

We can observe that the variables in general show some common pattern between the two datasets. The difference is acceptable as they are based on entirely different time frames. If we observe the model performance, it is evident that the model performance is weaker than obtined in the train dataset. The RMSE has increased by 383 from 1213 to 1596, however this is still 768 lower than the NULL model. The adjusted R-squared in our test dataset has also dropped by 12.3% to 62%. We also tested Random Forest model as an alternative of the OLS model, and althouth Random Forest showed promising results in our train dataset, the results were worse than OLS model for the test dataset.
